Research Notes

### 2018-04-12 writups/notes.tex
nonlin is working

Thus, if we impose BCs $\mathcal{P}(x, t) = \sin \frac{2\pi x}{L}$ and start
with initial conditions such that all quantities are zero, we would expect after
transients die out that
\begin{equation}
    \begin{split}
        P(x, z, t) &= \frac{\sin \frac{2\pi x}{L}}{\sinh 2\pi}
            \sinh \p*{2\pi \frac{L - z}{L}},\\
        u_{1x}(x, z, t) &= -\frac{2\pi t}{L\rho_0}
            \frac{\cos \frac{2\pi x}{L}}{\sinh 2\pi}\sinh \p*{2\pi \frac{L -
                z}{L}},\\
        u_{1z}(x, z, t) &= +\frac{2\pi t}{L\rho_0}
            \frac{\sin \frac{2\pi z}{L}}{\sinh 2\pi}\cosh \p*{2\pi \frac{L -
                z}{L}}.
    \end{split}\label{eq:incomp_nog_sol}
\end{equation}
This is in good agreement with the results, presented in \autoref{fig:no_g}.
Note that $P$ is constant while $\vec{u}$ increases linearly in time, and we
observe the expected $\sim \sin x \sinh \frac{L - z}{z}$ dependence. In fact,
$u_{1x}, u_{1z}$ are exactly $\frac{2\pi}{10}$ at $t = 1$.
\begin{figure}[!h]
    \centering
    \begin{subfigure}{0.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../sims/2d_no_g/no_g_P_t50.png}
    \end{subfigure}
    \begin{subfigure}{0.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../sims/2d_no_g/no_g_ux_t50.png}
    \end{subfigure}
    \begin{subfigure}{0.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../sims/2d_no_g/no_g_uz_t50.png}
    \end{subfigure}

    \begin{subfigure}{0.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../sims/2d_no_g/no_g_P_t100.png}
    \end{subfigure}
    \begin{subfigure}{0.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../sims/2d_no_g/no_g_ux_t100.png}
    \end{subfigure}
    \begin{subfigure}{0.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../sims/2d_no_g/no_g_uz_t100.png}
    \end{subfigure}
    \caption{$P, u_x, u_z$ at $t = 0.5$ and $t = 1$ for $\rho_0 = 1$. We choose
    a $L = 10$ square domain.}\label{fig:no_g}
\end{figure}

@@

We consider the problem where $\rho_1 \sim \rho_0$, where we cannot make such a
perturbative expansion. We write down thus nonlinear fluid equations
\begin{equation}
    \begin{split}
        \pd{\rho}{t} + \vec{u} \cdot \p*{\vec{\nabla}\rho} &= 0,\\
        \vec{\nabla} \cdot \vec{u} &= 0,\\
        \pd{\vec{u}}{t} &= -\frac{\vec{\nabla}P}{\rho} - g\hat{z}.
    \end{split}\label{eq:nonlin_incomp}
\end{equation}
Note that we cannot even subtract off hydrostatic equilibrium anymore since
$\rho$ can deviate greatly from the normal $\rho_0e^{-z/H}$!

Since Dedalus cannot handle a field quantity in the denominator, we must get rid
of $\rho$ somehow. We choose to do this by recasting the equations in terms of
$\rho^{-1}$, which is everywhere nonzero and finite since $\rho$ is everywhere
nonzero and finite. But then $\pd{\rho}{t} = -\rho^2\pd{\rho^{-1}}{t}$ while
$\vec{\nabla}\rho = -\rho^2\vec{\nabla}\rho^{-1}$, and so factoring out the
$\rho^2$ we can simply rewrite
\begin{equation}
    \begin{split}
        \pd{\rho^{-1}}{t} + \vec{u} \cdot \p*{\vec{\nabla}\rho^{-1}} &= 0,\\
        \vec{\nabla} \cdot \vec{u} &= 0,\\
        \pd{\vec{u}}{t} &= -\rho^{-1}\vec{\nabla}P - g\hat{z}.
    \end{split}\label{eq:nonlin_incomp}
\end{equation}

We must also consider how we implement the outgoing boundary condition in this
regime. If we use the radiative boundary condition formula, thankfully we have
little to change, though it also doesn't do a great job of supressing
reflections. If we instead use a sponge zone, we should be careful about how we
damp $\rho^{-1}$, since we obviously should not try damping it to zero! This is
why we initially chose to only apply our sponge zone to the velocity variables,
even in the linear problem.

@@

>>>>

Thus, if we impose BCs $\mathcal{P}(x, t) = \sin \frac{2\pi x}{L}$ and start
with initial conditions such that all quantities are zero, we would expect after
transients die out that
\begin{equation}
    \begin{split}
        P(x, z, t) &= \frac{\sin \frac{2\pi x}{L}}{\sinh 2\pi}
            \sinh \p*{2\pi \frac{L - z}{L}},\\
        u_{1x}(x, z, t) &= -\frac{2\pi t}{L\rho_0}
            \frac{\cos \frac{2\pi x}{L}}{\sinh 2\pi}\sinh \p*{2\pi \frac{L -
                z}{L}},\\
        u_{1z}(x, z, t) &= +\frac{2\pi t}{L\rho_0}
            \frac{\sin \frac{2\pi z}{L}}{\sinh 2\pi}\cosh \p*{2\pi \frac{L -
                z}{L}}.
    \end{split}\label{eq:incomp_nog_sol}
\end{equation}
This is in good agreement with the results, presented in \autoref{fig:no_g}.
Note that $P$ is constant while $\vec{u}$ increases linearly in time, and we
observe the expected $\sim \sin x \sinh \frac{L - z}{z}$ dependence. In fact,
$u_{1x}, u_{1z}$ are exactly $\frac{2\pi}{10}$ at $t = 1$.
\begin{figure}[!h]
    \centering
    \begin{subfigure}{0.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../sims/2d_0_no_g/no_g_P_t50.png}
    \end{subfigure}
    \begin{subfigure}{0.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../sims/2d_0_no_g/no_g_ux_t50.png}
    \end{subfigure}
    \begin{subfigure}{0.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../sims/2d_0_no_g/no_g_uz_t50.png}
    \end{subfigure}

    \begin{subfigure}{0.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../sims/2d_0_no_g/no_g_P_t100.png}
    \end{subfigure}
    \begin{subfigure}{0.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../sims/2d_0_no_g/no_g_ux_t100.png}
    \end{subfigure}
    \begin{subfigure}{0.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../sims/2d_0_no_g/no_g_uz_t100.png}
    \end{subfigure}
    \caption{$P, u_x, u_z$ at $t = 0.5$ and $t = 1$ for $\rho_0 = 1$. We choose
    a $L = 10$ square domain.}\label{fig:no_g}
\end{figure}

@@

We consider the problem where $\rho_1 \sim \rho_0$, where we cannot make such a
perturbative expansion. We write down thus nonlinear fluid equations
\begin{equation}
    \begin{split}
        \pd{\rho}{t} + \vec{u} \cdot \p*{\vec{\nabla}\rho} &= 0,\\
        \vec{\nabla} \cdot \vec{u} &= 0,\\
        \pd{\vec{u}}{t} &= -\p*{\vec{u} \cdot \vec{\nabla}} \vec{u}
            - \frac{\vec{\nabla}P}{\rho} - g\hat{z}.
    \end{split}\label{eq:nonlin_incomp}
\end{equation}
Note that we cannot even subtract off hydrostatic equilibrium anymore since
$\rho$ can deviate greatly from the normal $\rho_0e^{-z/H}$!

@@

\section{Conservative Fluid Equations}

This isn't particularly useful for our work since spectral methods don't benefit
from having the dynamical equations in conservative form, but form a useful
exercise for the ever-so-stupid writer.

The Reynolds transport theorem tells us that
\begin{equation}
    \rd{}{t}\int\limits_V Q\;\mathrm{d}V
        = \int\limits_V \pd{Q}{t}\;\mathrm{d}V
            + \int\limits_{\partial V}
                \p*{\vec{v}\p*{\mathrm{d}A} \cdot \hat{n}} Q\;\mathrm{d}A.
\end{equation}
We have notated $\vec{v}\p*{\mathrm{d}A}$ the velocity of the surface element.
In fluid mechanics, this is just the velocity field. Thus, for conserved
quantities where $\rd{}{t}\int\limits_V Q\;\mathrm{d}V = 0$, we can apply
divergence theorem to obtain the simple result
\begin{equation}
    \pd{Q}{t} + \vec{\nabla} \cdot \p*{Q\vec{v}} = 0.\label{eq:cons_eq}
\end{equation}
Of course, if there are any net forces etc.\ on the system, they are simply
inserted as fileds on the right hand side of \autoref{eq:cons_eq}.

The application of this to the three major conserved quantities: mass, momentum
and energy, yield:
\begin{description}
    \item[Conservation of Mass] This produces simply $\pd{\rho}{t} +
        \vec{\nabla} \cdot \p*{\rho \vec{v}} = 0$ the continuity equation.

    \item[Conservation of Momentum] This produces $\pd{\rho v_i}{t} +
        \vec{\nabla} \cdot \p*{\rho v_i \vec{v}} = 0$. To obtain the traditional
        momentum equation, it is perhaps most clear in index notation where
        repeated indicies denote summation:
        \begin{align}
            \pd{\p*{\rho v_i}}{t} + \partial_j \rho v_iv_j &= 0,\nonumber\\
            v_i \pd{\rho}{t} + \rho \pd{v_i}{t} + v_I \partial_j \p*{\rho v_j}
                + \rho v_j \partial_j v_i &= 0,\nonumber\\
            v_i\p*{\pd{\rho}{t} + \partial_j \p*{\rho v_j}}
                + \rho\p*{
                    \pd{v_i}{t} + v_j \partial_t v_i
                } = 0.
        \end{align}
        The first parenthetical term we recognize is just the continuity
        equation though. The remainder is the traditional source-free momentum
        equation.

    \item[Conservation of Energy] This just reads $\pd{e}{t} + \vec{\nabla}
        \cdot \p*{e \vec{v}} = 0$. It's a pretty uninteresting equation without
        pressure and thermodynamic effects, which would arise from a more
        careful treatment of the source terms.

        In fact, pressure contributes a source term $P\vec{\nabla} \cdot
        \vec{v}$, so the full energy equation is often written
        \begin{equation}
            \rd{e}{t} + \p*{e + P}\vec{\nabla} \cdot \vec{v} = 0.
        \end{equation}
\end{description}
